---
title: Preparation for Lab on Regulating Greenhouse Gas Emissions
lab_number: 11
lab_date: '2018-11-05'
pubdate: '2018-07-31'
date: '2018-11-05'
pdf_url: /files/lab_docs/regulation_lab/preparation_for_regulation_lab.pdf
slug: lab_11_preparation
output.blogdown::html_page:
  md_extensions: +tex_math_single_backslash+compact_definition_lists
  toc: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(lubridate)
# library(xtable)
library(knitr)
library(kableExtra)

game_day = ymd("2018-11-05")
long_game_day = format(game_day, "%A %B %e")
short_game_day = format(game_day, "%b.\ %e")

```
# Introduction

For the laboratory period on `r long_game_day`, we will do a role-playing 
exercise explore the effectiveness of three different approaches to 
regulating pollution, such as greenhouse gas emissions: command-and-control, 
cap-and-trade, and emissions taxes. 
This reading should serve to prepare you for the exercise.

# Zero-Emissions Baseline

The handout "The economics of regulating greenhouse gas emissions" uses a 
baseline scenario of business-as-usual emissions and calculates the costs and 
benefits of an alternate policy that cuts emissions below this baseline. 
This is the usual way policy analysts examine regulations, but if we did the 
emissions-trading game this way, it would be more confusing. To simplify things, 
we'll do the emissions-trading game a different way: with a zero-emissions 
baseline that will allow us to calculate costs and benefits of emitting CO~2~. 
In this case, the costs to the players will be the cost of buying permits to 
emit CO~2~ and the benefits will be the profits they earn. The costs to society 
will be the social cost of carbon (the damage caused by global warming) and the 
benefit will be the combination of profit earned by the companies (this benefits 
the companies, but it's also part of the gross domestic product of the nation) 
and money received from the players as taxes or payment for emissions permits.

# The Exercise



## The Players

To keep groups small, I will divide the class in half and each half will perform 
the same role-playing exercise. For the exercise, I will divide the students 
into three groups:

1. One group will play The Environmental Protections Agency (EPA) and will 
   decide how much to reduce pollution, how many permits to issue, or how 
   much to charge as a pollution tax. 
   
    The EPA's motivation is to produce the best net benefit for society by 
    balancing the costs of reducing greenhouse gas emissions against the 
    benefits of limiting global warming.
2. A second group will play Alpha Electricity: a large power company with a 
   varied portfolio of generating plants including coal, natural gas, and 
   nuclear. 
   
     Alpha's motives are purely to produce the greatest profit for its 
     shareholders regardless of the cost to society of greenhouse gas emissions.
3. A third group will play Beta Industries: a large heavy-industrial 
   conglomerate with many large factories producing steel, aluminum, and 
   petrochemicals such as plastics, paints, and pharmaceuticals. 
   
     Beta's motives are purely to produce the greatest profit for its 
     shareholders regardless of the cost to society of greenhouse gas 
     emissions.

For the purposes of this exercise, we will assume that the EPA can accurately 
estimate the damage that would be caused by global warming, and thus, that it 
can also accurately estimate the social benefit of reducing greenhouse gas 
emissions. However, the EPA cannot accurately assess the costs individual 
companies will incur when they reduce their emissions. This means that the 
EPA's estimates of net benefits (benefits minus costs) is limited by its 
uncertainty about the cost of reducing emissions.

Only the EPA will know the social cost of greenhouse gas emissions, only Alpha 
will know Alpha's cost for reducing emissions, and only Beta will know Beta's 
cost for reducing emissions.

Without regulation, Alpha and Beta would each emit 150 million tons of CO~2~ 
per year. The goal of the exercise is for the EPA to reduce pollution to achieve 
the socially optimal balance between the benefits of economic activity 
(jobs, wealth, etc.) and the harms of pollution. The goal of each firm is to 
maximize their profit regardless of the costs or benefits to society or to its 
competitor.

To keep things simple, emissions cuts will be figured in blocks of 
10 million tons, so a firm can cut emissions by zero, 10 million tons, 
20 million tons, ..., up to a maximum of 150 million tons (cutting emissions 
by 150 million tons means the firm reduces its emissions to zero).

\clearpage
## The Game

The game will have six stages:

1. First, the EPA will gather information on the cost of abating pollution. 
   A representative of the EPA can ask four questions about each company's 
   costs (I recommend asking questions about the marginal profit for the 
   $n^{\text{th}}$ million tons of pollution, or how much it would cost to 
   reduce emissions by $x$ million tons.). A representative of each company 
   will answer the question. The representatives may answer strategically, 
   meaning they may exaggerate the costs. The EPA may take this into account 
   in deciding how to use the answers to estimate the true cost of reducing 
   emissions.
2. Second, the EPA will determine three possible courses of action:
    a) A command and control regulation, which mandates a specific emissions 
       reduction. Because the Constitution guarantees equality before the law, 
       this regulation must impose the same emissions cut for each 
       firm.^[The real world is a bit more complex, because the government 
       would be allowed to regulate different industries differently, but would 
       face real obstacles if it tried to impose different emissions cuts on 
       different companies within the same industry. Because this exercise is 
       limited to just two companies, we simplify by forcing the EPA to impose 
       identical emissions cuts on all firms.} Based on what it knows about the 
       costs (to the firms) and benefits (to society) of reducing emissions, the 
       EPA will determine how many million tons of total emissions to cut 
       (to keep things simple, the EPA should make the amount an even multiple 
       of 20 million tons). It will then divide those emissions cuts equally 
       between the two firms.]
    b) A cap-and-trade. Under this program, the total emissions cuts would be 
       the same and the EPA will issue permits to emit CO~2~. Each permit will 
       allow the owner to emit 10 tons of CO~2~. Total CO~2~ emissions are 
       300 million tons minus the emissions cuts the EPA wants to impose. 
       If the EPA wants to cut emissions by 200 million tons, it would issue 
       10 permits 
       ($300~\text{million} - 20 \times 10~\text{million} = 100~\text{million}$; 
       Each permit allows 10 million tons of emissions so 
       $100~\text{million}/10~\text{million} = 10~\text{permits}$). 
       Again, to keep things simple, make the number of permits a multiple of 2. 
       The EPA will give each company an equal number of permits (if the EPA 
       issues 10 permits, it gives 5 to each company).
	  c) An emissions-tax program. Under this program, the EPA determines the tax 
		   a firm must pay for every million tons of CO~2~ it emits. The firms will 
		   then decide on their own how much to cut their emissions.
3. Third, the firms determine how much CO~2~ they will emit under the 
   command-and-control program (this is easy). At the end of this round, the 
   EPA will publish the total benefit to society of the emissions reduction, 
   the firms will calculate their total costs (but keep these secret), and the 
   instructor will announce the total deadweight loss due to inefficiencies in 
   the regulation.
4. Fourth, permits will be distributed. The class will vote whether to auction 
   the permits and distribute them to the highest payer or give equal numbers 
   of permits to each firm and allow the firms to trade permits with each other. 
   After the permits are allocated, we will again announce the total benefits to 
   society and the deadweight loss.
5. Fifth, we will impose an carbon tax, at the price set by the EPA in stage 2. 
   The two firms will be free to cut emissions by as much or as little as they 
   want (between 0 and 150 million tons each). After each firm determines its 
   final emissions, we will once again announce the total benefits to society 
   and the deadweight loss.
6. Sixth and finally, we will reveal the details at each step, including the 
   private information each firm has about its costs. This will let us discuss 
   and analyze the strengths and limitations of each regulatory program.

## Acknowledgements

This exercise was adapted from The Pollution Game, an interactive exercise 
developed by Jay R. Corrigan, Associate Professor of Economics, Kenyon College.  
See J.R. Corrigan, ``The Pollution Game: A Classroom Exercise Demonstrating the 
Relative Effectiveness of Emissions Taxes and Tradable Permits,'' 
_The Journal of Economic Education_ **42**, 70--78 (2011) 
[doi: 10.1080/00220485.2011.536491](https://doi.org/10.1080/00220485.2011.536491)

\clearpage
## Homework

Do the following homework (turn it in on `r long_game_day` at the beginning of 
lab):

1. Table 1 lists the marginal profit a company earns by emitting CO~2~. 
   Fill in the blanks in the table to show the total profits it earns at each 
   amount of CO~2~ emissions.

2. What level of CO~2~ emissions would produce the maximum total profit? 
   How much profit would this be?
   \vspace{2cm}

3. If the company is emitting the amount of CO~2~ that would maximize its 
   profits, and then the Environmental Protection Agency requires the company 
   to reduce its emissions by 5 million tons, what is the total cost for the 
   company to comply with the regulation?
   \vspace{2cm}

4. What is the marginal cost to comply with the regulation? 
   (Be careful and consider, if the company cuts 1 million tons, then a second 
   million tons, and so forth, what did it cost the company to make the 
   fifth million-ton cut?)
   \vspace{2cm}

5. If the EPA imposed a tax of \$30 per ton on CO~2~ emissions, complete the 
   table to indicate the new marginal profit and total profit at each level of 
   emission.
   \vspace{2cm}

6. Under a \$30 per ton tax, what level of CO~2~ emissions would produce the 
   maximum total profit? What would its total profit be? How much less is this 
   than the total profit you reported in question 2?
   \vspace{2cm}

7. Table 2 shows the economic value of the marginal 
   environmental harm caused by each additional million tons of CO~2~ emissions. 
   Fill in the blanks to indicate the total environmental harm.

8. The marginal net economic impact is the marginal profit generated by emitting 
   each million tons of CO~2~ minus the marginal environmental harm. This number 
   is the net benefit to society from emitting an additional million tons of 
   CO~2~. If the number is positive, society benefits. If it is negative, society 
   suffers.

    Fill in the marginal and total economic impacts, using the information on 
    environmental harm from Table 2 and the information 
    on marginal profits from Table 1.

9. What is the optimum amount of CO~2~ to emit if we consider the net benefit to 
   society (i.e., the net economic impact).
   \vspace{2cm}

10. If you were going to set a cap on emissions, how many tons would you set the 
    cap at?
   \vspace{2cm}

11. If you were going to set a tax on emissions, how many dollars per ton would 
    you set the tax at?
   \vspace{2cm}

### Table 1: Profits vs. Emissions

```{r profit_table}
tax = 30
profits = tibble(
  emissions = seq(1,20), 
  marginal_profit = 91 - 6 * emissions,
  total_profit = str_c("$", marginal_profit, 
                       ifelse(is.na(lag(marginal_profit)), "", 
                              str_c(" + $", lag(cumsum(marginal_profit)), 
                                    " = $", cumsum(marginal_profit))
                              )
                       ),
  marginal_taxed = str_c("$", marginal_profit, " - $", tax, " = $",
                         marginal_profit - tax),
  total_taxed = str_c("$", marginal_profit - 30,
                       ifelse(is.na(lag(marginal_profit)), "", 
                              str_c(" + $", lag(cumsum(marginal_profit - 30)), 
                                    " = $", cumsum(marginal_profit - 30))
                              )
                      ),
) %>% mutate_at(vars(total_profit:total_taxed),
                funs(ifelse(emissions <= 3, ., "")))

column_names <- c("CO~2~ Emissions\n(million tons)",
                    "Marginal profit\n(million dollars)", 
                    "Total profit\n(million dollars)",
                    "Marginal profit\n(million dollars)", 
                    "Total profit\n(million dollars)")

output_format <- ifelse(is_html_output(), "html", "latex")
# message("Output format = ", output_format)

if (output_format == "latex") {
   escape = FALSE
   column_names <- linebreak(column_names) %>%
      str_replace_all("CO~2~", "$\\\\text{CO}_2$")
   profits <- profits %>% mutate_all(funs(str_replace_all(., fixed("$"), "\\$")))
} else {
   lb <- function(x) x
   escape = TRUE
}

kable(profits, format = output_format, align = "crrrr", digits = 0, 
      col.names = column_names,
      booktabs = T, escape = FALSE) %>% 
  add_header_above(c(" " = 1, "No Emissions Tax" = 2, 
                     "$30/ton Emissions Tax" = 2)) %>%
  kable_styling(c("bordered","hover"), c("striped", "scale_down"))
```

**Table 1:** Profits versus emissions: The table lists the marginal profit, in 
millions of dollars, for each million tons of CO~2~ it emits. For instance, if 
it emits one million tons, it earns a marginal profit of $85 million. 
The total profit for _x_ million tons is the sum of the marginal profits for 
each million tons from 1 to _x_.

Fill in the blanks with the total profit the company earns by for each amount 
of CO~2~ emission."

### Table 2: Environmental Harm vs. Emissions

```{r cost_table}
costs = tibble(
  emissions = seq(1,20), 
  marginal_profit = 91 - 6 * emissions,
  marginal_harm = 5 * emissions,
  impact = marginal_profit - marginal_harm,
  total_harm = str_c("$", marginal_harm, 
                       ifelse(is.na(lag(marginal_harm)), "", 
                              str_c(" + $", lag(cumsum(marginal_harm)), 
                                    " = $", cumsum(marginal_harm))
                              )
                       ),
  marginal_impact = str_c("$", marginal_profit, " - $", marginal_harm, 
                          " = $", impact),
  total_impact = str_c("$", cumsum(impact))
) %>% mutate_at(vars(total_harm:total_impact),
                funs(ifelse(emissions <= 3, ., ""))) %>%
  select(-marginal_profit, -impact)

column_names <- c("CO~2~ Emissions\n(million tons)",
                    "Marginal harm\n(million dollars)", 
                    "Total harm\n(million dollars)",
                    "Marginal impact\n(million dollars)", 
                    "Total impact\n(million dollars)")

output_format <- ifelse(is_html_output(), "html", "latex")
# message("Output format = ", output_format)

if (output_format == "latex") {
   column_names <- linebreak(column_names) %>%
      str_replace_all("CO~2~", "$\\\\text{CO}_2$")
   escape = FALSE
   costs <- costs %>% mutate_all(funs(str_replace_all(., fixed("$"), "\\$")))
} else {
   escape = TRUE
}


kable(costs, format = output_format, align = "crrrr", digits = 0,
      col.names = column_names, booktabs = T, escape = escape) %>% 
  add_header_above(c(" " = 1, "Environmental Harm" = 2, 
                     "Net Economic Impact" = 2)) %>%
  kable_styling(c("bordered","hover"), c("striped", "scale_down"))
```

**Table 2**: Economic impacts of CO~2~ emissions. The table lists the marginal 
environmental harm, in millions of dollars, for each million tons of CO~2~ 
released into the environment. For instance, the first million tons of 
emissions cause $5 million in damage.

The right pair of columns shows the net economic impact of the profits 
generated from the emissions (see Table 1) minus the environmental harm.

Fill in the blanks with the total environmental harm from CO~2~ emission and 
compute the net economic impact.
