---
title: Instructions for Top-Down Decarbonization Policy Analysis
lab_number: 10
lab_date: '2018-10-29'
pubdate: '2018-07-31'
date: '2018-10-29'
pdf_url: /files/lab_docs/decarb_lab_top_down/decarb_top_down_instructions.pdf
slug: lab_10_instructions
output.blogdown::html_page:
  md_extensions: +tex_math_single_backslash+compact_definition_lists
  toc: yes
---
```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)

is_output_html <- function() {
  is_html <- TRUE
  header <- rmarkdown:::parse_yaml_front_matter(
    readLines(knitr::current_input())
  )
  if ("output.blogdown::html_page" %in% names(header)) {
    is_html <-  TRUE
  } else if ("output" %in% names(header)) {
    if (is.list(header$output)) {
      output <- names(header$output)[1]
    } else {
      output <- header$output[1]
    }
    if (str_detect(output, regex("(pdf|word)_document", ignore_case = TRUE))) {
      is_html <- FALSE
    }
  }
  is_html
}


getOutputFormat <- function() {
  output <- rmarkdown:::parse_yaml_front_matter(
    readLines(knitr::current_input())
    )$output
  if (is.list(output)){
    return(names(output)[1])
  } else {
    return(output[1])
  }
}

data_dir = "_data"
script_dir = "_scripts"

library(pacman)
p_load(magrittr, tidyverse, lubridate, knitr, broom)
p_load_gh("jonathan-g/kayadata")

theme_set(theme_bw(base_size = 15))

if (!dir.exists(data_dir)) dir.create(data_dir)

source(file.path(script_dir, "format_md.R"), chdir = T)

due_date_1 <- date("2018-10-30")
due_date_2 <- date("2018-11-12")

long_due_date_1 <- stamp("Monday, March 1")(due_date_1)
short_due_date_1 <- stamp("Mar. 1")(due_date_1)

long_due_date_2 <- stamp("Monday, March 1")(due_date_2)
short_due_date_2 <- stamp("Mar. 1")(due_date_2)

start_year <- max(get_kaya_data("United States")$year, na.rm = T)
after_start_year <- start_year + 1

eia_report_year <- max(get_top_down_values("United States")$year, na.rm = T)
donut_year <- max(kayadata::get_fuel_mix("United States")$year, na.rm = T)

IEO_url <- "http://www.eia.gov/forecasts/ieo/"
EIA_table_url <- "http://www.eia.gov/oiaf/aeo/tablebrowser"

top_down_country_assignments <- bind_rows(
  tibble(class = "Undergraduates", 
         region = c("World", "United States", "China")),
  tibble(class = "Grad Students", 
         region = c("World", "United States", "China",
                     "India", "Brazil"))
)

rcp_26 <- tribble(
  ~region, ~target_year, ~ref_year, ~reduction,
  "Australia/New Zealand", 2050, 2005,  0.82,
  "Canada",                2050, 2005,  0.72,
  "China",                 2050, 2005,  0.78,
  "India",                 2050, 2005,  0.73,
  "Japan",                 2050, 2005,  0.66,
  "South Korea",           2050, 2005,  0.67,
  "United States",         2050, 2005,  0.73,
  "Africa",                2050, 2005,  0.28,
  "Latin America",         2050, 2005,  0.40,
  "Middle East",           2050, 2005,  0.32,
  "Southeast Asia",        2050, 2005, -0.17,
  "Western Europe",        2050, 2005,  0.74,
  "World",                 2050, 2005,  0.36
)

target_year <- rcp_26$target_year[1]
ref_year <- rcp_26$ref_year[1]
current_year <- get_fuel_mix("World")$year[1]
```
# Introduction

In this lab, you will use a top-down approach to figure out how much new energy 
infrastructure you would need to install for each country or region in order
to meet the emissions-reduction goals for `r target_year`.


You will analyze the same countries and regions you did in the bottom-up lab last 
week:
```{r country_assignments, echo=FALSE}
top_down_country_assignments %>% group_by(class) %>%
  summarize(region = str_c(region, collapse = ", ")) %>%
  ungroup() %>% 
  mutate(class = ordered(class, 
                         levels = c("Undergraduates", "Grad Students"))) %>%
  arrange(class) %>%
  kable()
```

The lab report is due (knitted to PDF or Word document, committed, and pushed
to GitHub) by Monday Nov.\ 12.

# The Assignment:

## Outline:

1. For each country, examine the top-down projections for the Kaya 
   variables for the year `r target_year`. How many quads of energy does the 
   Energy Information Administration preduct that the country will use
   in `r target_year`?
2. Examine the mix of energy sources that the country used in `r current_year`. 
3. If the  country uses the same percentages of each energy source in 
   `r target_year` that it did in `r current_year`, 
   calculate how much energy the country would use from each energy source in
   `r target_year` and how much CO~2~ each energy source would emit.
   Add these up to get the total CO~2~ emissions _F_ in `r target_year` under a
   "business-as-usual" scenario.
4. Calculate the policy target for _F_ in the year `r target_year`.
5. Calculate how much the country would have to reduce its CO~2~ emissions  in 
   `r target_year` below business-as-usual to meet its policy goal.
6. Allocate the emissions reduction from step 5 across the different fuels and 
   calculate the amount of fossil fuel energy the country would need to replace 
   with clean energy by `r target_year` to meet its policy goal.
7. Calculate how many clean power plants the country would need to build 
   between now and `r target_year` to produce this clean energy. 
   
    Report this as the total number, the number per year, and  the number per 
    week or per day if appropriate.
  
    Start by doing this calculation for nuclear energy plants, then repeat the 
    calculation for solar, and then for wind.
8. Suppose instead that the country or region supplied its growing energy 
   demand with coal. Calculate how many new coal power plants it would have
   to build between now and `r target_year` to supply the growing energy 
   demand. 
   
    Calculate the number of new coal power plants the country would have to 
    build and how much additional CO~2~ that would put into the atmosphere
    every year.

## Detailed Instructions:

For each country:

1. Use the function `get_top_down_values()` from the 
   `kayadata` package to examine the top-down projections for the Kaya 
   variables for the year `r target_year`. How many quads of energy does the 
   Energy Information Administration preduct that the country will use
   in `r target_year`?
2. Use the `get_fuel_mix()` function to get the mix of energy sources that the
   country used in `r current_year`. 
3. If the  country uses the same percentages of each energy source in 
   `r target_year` that it did in `r current_year`, calculate the number of
   quads and the greenhouse gas emissions from each energy source 
   in `r target_year`, if the total energy consumed in that year
   is the value for _E_ that you looked up in step (1).
   
    First, multiply the total energy demand _E_ in `r target_year` by the 
    percentages in the fuel mix to get the number of quads for that energy
    source. Then multiply the number of quads for each energy source by the 
    emissions factor for that fuel to get the emissions from that energy
    source.
    
    Finally, add up the emissions from all of the fuel sources to get the total
    emissions for `r target_year`, under a "business-as-usual" scenario, where
    the mix of energy sources does not change.
    
    **Hint:** You can combine data frames using `join` functions to make this
    easier. Here is an example, which I have worked for Mexico. You can
    do the same thing for other countries or regions:
    ```{r example_fuel_mix}
    E_2050 <- get_top_down_values("Mexico") %>% 
      filter(year == 2050) %$% E
    fm <- get_fuel_mix("Mexico") %>% select(-region_code, -geography)
    # Calculate the mix of energy sources for 2050
    fm_2050 <- fm %>% mutate(quads = E_2050 * pct)
    kable(fm, digits = 2)
    ```
    
    ```{r example_fuel_mix_2}
    # Get the emission factors
    factors <- emissions_factors()
    kable(factors, digits = 2)
    ```
    
    ```{r example_fuel_mix_3}
    # Combine the emission factors data frame with the 
    # fuel mix data frame:
    fm_2050 <- fm_2050 %>% left_join(factors, by = "fuel") %>%
      # This line is to fix a problem with the kaya_data package,
      # because I forgot to put an emissions factor for Hydro power.
      mutate(emission_factor = replace_na(emission_factor, 0))
    kable(fm_2050, digits = 2)
    ```
    
    ```{r example_fuel_mix_4}
    fm_2050 <- fm_2050 %>% mutate(emissions = quads * emission_factor)
    
    # Summarize by adding up the emissions from each fuel source.
    total_emissions_2050 <- fm_2050 %>% 
      summarize(emissions = sum(emissions)) %$% emissions
    ```
    
4. Get the Kaya identity data for the country from `get_kaya_data()` and look
   up the CO~2~ emissions _F_ in `r ref_year`. Then use the policy goal for 
   emissions reduction from the data frame `rcp_26` to calculate the policy 
   target for _F_ in the year `r target_year`. You did this calculation last
   week as part of the bottom-up analysis.
5. Compare the projected "business as usual" emissions from step (3) to the 
   policy target from step (4) to figure the total amount of emissions you 
   would need to cut by converting fossil fuel energy sources to clean sources.
6. Use the top-down procedure I presented in class last week to allocate the
   emissions reduction that you calculated in step (5) across the different 
   energy sources for `r target_year`, that you calculated in step (3). 
   
    How many million metric tons of CO~2~ would you cut from coal, natural gas, 
    and oil? How many quads of energy from that fuel would you need to replace
    with new clean energy?
    
    Add up all of the clean energy requirements to calculate the number of quads
    of clean energy that the country would need to add between now and 
    `r target_year` to meet the policy goal.
7. Calculate how many clean generating plants you would need to build between
   now and `r target_year` to produce this clean energy. Report this as the
   total number, the number per year, and  the number per week or per day if
   the number is large.
  
    Start with nuclear, then do the calculation for solar and then for wind.
    
    You can look up the nameplate capacity and the capacity factor for 
    different kinds of power plants (clean and dirty) from the function 
    `generating_capacity()`.  
    Remember that the average number of megawatts a power source supplies over 
    a year is the nameplate capacity times the capacity factor.
    Remember that one quad equals 
    `r format_md(megawatts_per_quad(), comma = TRUE)` megawatts.
8. Suppose instead that the country or region supplied its growing energy 
   demand with coal. Calculate how many new coal power plants it would have
   to build between now and `r target_year` to supply the growing energy 
   demand. 
   
    Calculate the number of new coal power plants the country would have to 
    build and how much additional CO~2~ that would put into the atmosphere
    every year.
