---
title: 'Project: Decarbonization Policy for One Country'
lab_number: 12
lab_date: '2018-11-12'
pubdate: '2018-07-31'
date: '2018-11-12'
bibliography: decarbonization.bib
pdf_url: /files/lab_docs/decarb_project/country_decarbonization_assignment.pdf
slug: lab_12_assignment
output.blogdown::html_page:
  md_extensions: +tex_math_single_backslash+compact_definition_lists
  toc: yes
---
```{r setup, include=FALSE, cache = FALSE}
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(cache = FALSE, echo = TRUE, fig.height=4, fig.width=7)

# This section loads necessary R libraries and sources scripts that define
# useful functions format_md.
#
data_dir = "_data"
script_dir = "_scripts"

if (!dir.exists('data')) dir.create('data')

library(pacman)

# Load all of the following packages.
# Install any missing packages.
p_load(scales, tidyverse, stringr, janitor, knitr)
p_load_gh("jonathan-g/kayadata")

# This sets the default style for ggplot
theme_set(theme_bw(base_size = 15))

# Load basic utility scripts
source('_scripts/format_md.R', chdir = T)
```
# Introduction

This assignment will be more free-form than the lab assignments we have done
earlier in the semester.
For this assignment, you will build on the work you did for the bottom-up and
top-down decarbonization analysis and investigate how decarbonization might
play out in one specific country.
You can work on your own or in a team of two or three for this assignment.
You will choose one country, decide what that country's greenhouse gas
emissions target should be for the year 2050, and then sketch out a plan for
how that country could achieve its goal.

You and your team will give a short (five minutes per person) presentation in
the final lab on Monday, December 3 and turn in a written report (one report
from the whole team is acceptable)
before midnight on Wednesday December 5.
The report is not meant to be an exhaustive research report. I expect a length
of around 5 pages of doublespaced
text (or equivalently, around 3 pages singlespaced)
per team
member, plus appropriate figures, tables, and references.

Unlike previous assignments, you may use RMarkdown, Word, or any other
word-processing mode that you wish when writing your report, but I would like
to you produce PDF output (e.g., if you're writing in Word, save a copy in PDF
format).

# Establishing a Goal

Briefly, look at what your country has pledged to do for emissions reduction
for the 2015 Paris Agreements
(<http://spappssecext.worldbank.org/sites/indc/Pages/INDCHome.aspx> and the
individual Country Briefs at 
<http://spappssecext.worldbank.org/sites/indc/Pages/Content_Brief.aspx>).
Consider what the country is planning to do by 2030 or whichever alternative
target year it specifies in its Nationally Determined Contribution and
then consider how much more the country might reasonably plan to do by
2050 to make a fair contribution to reducing global greenhouse gas emissions
while also taking into account the need for less-developed nations to
grow their economies in order to lift people out of poverty.

For simplicity, I recommend that you focus on carbon dioxide emissions from
burning fossil fuels rather than on all different kinds of greenhouse gases.

There is no right answer, and it would be possible to write a 50-page
paper just on this part, but that is not what I want you to do.
Rather, just present a simple overview of the issues your country faces and how
you would choose a goal.

For instance, you could look back to 1965 (when the data on energy and emissions
in the `kayadata` package begins) and add up the total emissions from your
country and compare it to the total emissions of the world during that period
and then consider how your country's fraction of total emissions compare to your
country's fraction of the total world population or the total world Gross
Domestic Product (GDP).

You may want to refer to the table below, which lists emissions reductions from
2005--2050 for different parts of the world that would meet a global 36%
emissions target in 2050. This table comes from the IPCC's representative
concentration pathway (RCP) database and is based on the pathway called
"RCP 2.6,"" which gives about a two-thirds probability of keeping global warming
below 2&deg; C.
However, this table was produced from an economic model and does not take
account of political and ethical considerations, such as fairness, so you are
not obliged to choose the same goal that this table lists for your country or
its region.

```{r rcp_26, echo=FALSE}
rcp_26 <- tribble(
  ~region, ~target_year, ~ref_year, ~reduction,
  "Australia/New Zealand", 2050, 2005,  0.82,
  "Canada",                2050, 2005,  0.72,
  "China",                 2050, 2005,  0.78,
  "India",                 2050, 2005,  0.73,
  "Japan",                 2050, 2005,  0.66,
  "South Korea",           2050, 2005,  0.67,
  "United States",         2050, 2005,  0.73,
  "Africa",                2050, 2005,  0.28,
  "Latin America",         2050, 2005,  0.40,
  "Middle East",           2050, 2005,  0.32,
  "Southeast Asia",        2050, 2005, -0.17,
  "Western Europe",        2050, 2005,  0.74,
  "World",                 2050, 2005,  0.36
)

rcp_26 %>% mutate(reduction = percent(reduction, 1)) %>% select(-ref_year) %>%
                  kable(digits = 0, align = c("c", "r", "r"),
                        col.names = c("Region", "Year",
                                      str_c("Reduction from ", rcp_26$ref_year[1])))
```

# Realizing the Goal

Once you have chosen a goal, you should apply the same kinds of methods that we
used in the bottom-up and top-down decarbonization analyses to estimate what
your conuntry's population and per-capita GDP may be in 2030 and 2050, and what
the implications are for energy efficiency and the mixture of fuels that should
supply its energy needs. This is a place to research the country's natural
resources, current energy supply, and opportunities to decarbonize.

For instance, if your country is mountainous with many rivers, you may want to
expand hydroelectricity.
On the other hand, if it is relatively flat, has few rivers, or if all the
major rivers already have dams and generators, then you may want to look at
other sources of energy.
If your country receives a lot of sunshine, you may want to emphasize solar
energy.
Don't forget to think about nuclear energy as well as renewables like
hydroelectricity, wind, and solar.

I am not asking you to give a thorough engineering and economic assessment of
the energy transition, but to give a brief overview of the major opportunities
and obstacles to clean energy in your country and what you think would be the
best strategy for reducing emissions.

# Format

This is a lab report, not a formal research paper, so you do not need to
structure it as a formal paper.
My grading rubric for the written report will be:

* Thoughtful and sensible analysis of goals [30%]
* Thoughtful and sensible analysis of how to realize the goal [30%]
* Good use of the data for your country [20%]
* Organization of the report (do the parts fit together well and tell a clear
  story) [10%]
* Quality of writing [5%]
* Appropriate use of citations and references [5%]

You may choose to turn this in as an RMarkdown document (knitted to PDF, of
course) or a Word document (saved as PDF), or any other format that you have
rendered as PDF.

The key is that you explain clearly how you do your analysis.
You can either use RMarkdown or you can turn in a document written with a
word-processor and refer the reader to R scripts or an RMarkdown document
with "supporting information" that contains the details of your calculations
and quantitative analysis.

You will turn this assignment in by accepting the assignment on
GitHub Classroom, cloning the repository to your own computer, saving your
final products in the repository and committing and pushing them to GitHub.

# Data Sources
The `kayadata` package has data on the Kaya identity variables for 81 countries
and top-down projections for 78 countries.

If you are working on a country that's not one of these, you may want to look
at additional data, such as the International Energy Agency's energy and
emissions analyses.
These reports are extremely expensive to buy, but the Vanderbilt Library 
provides free online access to the World Energy Outlook reports from 1999--2017 
through the ACORN library catalog and this link
<https://www.oecd-ilibrary.org/energy/world-energy-outlook-2017_weo-2017-en>
if you are on campus or this link if you are off-campus
<https://www-oecd-ilibrary-org.proxy.library.vanderbilt.edu/energy/world-energy-outlook-2017_weo-2017-en>.
You can also get detailed data for energy in the the OECD nations at
<https://www.oecd-ilibrary.org/energy/energy/indicator-group/english_379b6cdc-en>)
or 
<https://www-oecd-ilibrary-org.proxy.library.vanderbilt.edu/energy/energy/indicator-group/english_379b6cdc-en>).

The U.S. Energy Information Administration
publishes detailed country profiles for a number of countries at 
<https://www.eia.gov/outlooks/ieo/>,
with data on many individual countries at
<https://www.eia.gov/beta/international/>
and brief country and regional analysis reports at
<https://www.eia.gov/beta/international/analysis.cfm>

This list of data sources is not exhaustive and you should feel free to explore
other sources of information, but don't feel that you have to spend hours in the
library or the internet chasing down data. It is fine for this assignment to use
"good enough" data or to substitute estimates if you can't find exact numbers
for what you're looking for.
Just be clear about how you're doing your analysis.
