---
title: Answers
lab_number: 3
lab_date: '2018-09-10'
pubdate: '2018-09-21'
date: '2018-09-17'
pdf_url: /files/lab_solutions/lab_03/lab_03_answers.pdf
slug: lab_03_answers
output.blogdown::html_page:
  md_extensions: +tex_math_single_backslash+compact_definition_lists
  toc: yes
---
```{r setup, include=FALSE}
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, message=FALSE, warning=FALSE,
                      fig.height=4, fig.width=6,
                      fig.pos = 'htbp')

# This section loads necessary R libraries and sources scripts that define 
# useful functions format_md.
# 
data_dir = "_data"
script_dir = "_scripts"

if (!dir.exists(data_dir)) dir.create(data_dir)

library(pacman)
p_load(zoo, xml2, tidyverse, stringr)

theme_set(theme_bw(base_size = 15))

source(file.path(script_dir, "utils.R"), chdir = T)
source(file.path(script_dir, "format_md.R"), chdir = T)
source(file.path(script_dir, "modtran.R"), chdir = T)

```
# Chapter 4 Exercises

## Exercise 4.1: Methane

Methane has a current concentration of 1.7 ppm in the atmosphere and
is doubling at a faster rate than CO~2~.

a) Would an additional 10 ppm of methane in the atmosphere have a larger or smaller
  impact on the outgoing IR flux than an additional 10 ppm of CO~2~ at current
  concentrations?

**Answer:**
Run MODTRAN in three configurations: 
400 ppm CO~2~ and 1.7 ppm methane,
400 ppm CO~2~ and 11.7 ppm methane,
and
410 ppm CO~2~ and 1.7 ppm methane,
and compare I~out~ for each configuration:

```{r ex_4_1_a_modtran, include=TRUE, message=FALSE}
run_modtran(file.path(data_dir, "modtran_baseline.txt"), 
            atmosphere = 'tropical', co2 = 400, ch4 = 1.7)
run_modtran(file.path(data_dir, "modtran_plus_10_ppm_methane.txt"), 
            atmosphere = 'tropical', co2 = 400, ch4 = 11.7 )
run_modtran(file.path(data_dir, "modtran_plus_10_ppm_co2.txt"), 
            atmosphere = 'tropical', co2 = 410, ch4 = 1.7 )

baseline = read_modtran(file.path(data_dir, "modtran_baseline.txt"))
plus_10_methane = read_modtran(file.path(data_dir, "modtran_plus_10_ppm_methane.txt"))
plus_10_co2 = read_modtran(file.path(data_dir, "modtran_plus_10_ppm_co2.txt"))

i_baseline = baseline$i_out
i_methane = plus_10_methane$i_out
i_co2 = plus_10_co2$i_out
```

Now, we calculate the change in I~out~ corresponding to adding 10 ppm of methane
and adding 10 ppm CO~2~:

```{r ex_4_1_a_delta_i_out, include=TRUE}
delta_i_methane = i_baseline - i_methane
delta_i_co2 = i_baseline - i_co2
ratio_10_ppm = delta_i_methane / delta_i_co2
```

At the default settings, 
I~out~ = `r format_md(i_baseline, digits = 2)` W/m^2^. 
If we add 10 ppm methane, I~out~ becomes
`r format_md(i_methane, digits = 2)` W/m^2^
(a decrease of `r format_md(delta_i_methane, digits = 2)` 
W/m^2^)
and if we add 10 ppm CO~2~, I~out~ becomes
`r format_md(i_co2, digits = 2)`
(a decrease of `r format_md(delta_i_co2, digits = 2)` 
W/m^2^). 

Thus, the impact of adding 10 ppm of methane on I~out~ is 
`r format_md(ratio_10_ppm, digits = 1)` 
times greater than the impact of adding 10 ppm of CO~2~,
which means methane would also have a much greater impact on the temperature.

b) Where in the spectrum does methane absorb? What concentration does it take to
  begin to saturate the absorption in this band? Explain what you are looking 
  at to judge when the gas is saturated.
  
    **Note:** See the suggestions in the instructions for this lab.

**Answer:**
We can run MODTRAN with no greenhouse gases except methane to find where it
absorbs. Setting methane to 10 ppm will give a good indication of where it
absorbs:

```{r ex_4_1_b_spectrum, include=TRUE, message=FALSE, warning=FALSE}
# file.path combines one or more directories with a filename.
filename = file.path(data_dir, "modtran_10_ppm_methane.txt")

run_modtran(filename, co2_ppm = 0, ch4_ppm = 10, 
            trop_o3_ppb = 0, strat_o3_scale = 0, h2o_scale = 0, 
            freon_scale = 0, 
            delta_t = 0, h2o_fixed = "vapor pressure",
            atmosphere = "tropical", clouds = "none", 
            altitude_km = 70, looking = "down")

plot_modtran(filename)
```

Methane absorbs in the range 1200--1400 cm^-1^ (we can look at the wavelength
scale on the top of the plot and see that this corresponds to roughly 
7--8 micron wavelength).

Now let's find out what concentration of methane band corresponds to the onset
of band-saturation. Start with 0.4 ppm methane and keep doubling for 11 
doublings (until we get to 2048 times the original concentration).

We create an empty `tibble`, and then every time we run a new
MODTRAN simulation, we create a tibble with one row and two columns:
methane concentration and I~out~. Then we use `bind_rows` to add this
tibble to the bottom of `methane_data`. This produces a tibble with
columns for methane concentration and I~out~, and a row for each run
of MODTRAN.

This is an example of using `bind_rows` to combine tibbles together
row by row when they have the same columns.

```{r ex_4_1_b_saturation, include=TRUE, message=FALSE, warning=FALSE}
        methane_data = tibble() # create a blank data tibble

        # The for command repeats everything between the braces "{...}"
        # for each value of x in the sequence 0, 1, 2, ..., 10, 11.    
        for (x in 0:11) {
          # Set the methane concentration to 0.4 times 2 to the power of x,
          # In other words 0.4, 0.8, 1.6, 3.2, ...
          p_methane = 0.4 * (2^x) 
          
          # Create a character variable that will be a file name of the form
          # "_data/methane_xx_x.txt", where xx_x is the methane concentration,
          # with an underscore for the decimal point.
          file_name = formatC(p_methane, digits = 1, decimal.mark = "_", 
                              format = "f") %>%
                      str_c('methane_', ., ".txt") %>%
                      file.path(data_dir, .)
    
          # Now run MODTRAN
          run_modtran(file_name, co2_ppm = 0, ch4_ppm = p_methane, 
                      trop_o3_ppb = 0, strat_o3_scale = 0, h2o_scale = 0, 
                      freon_scale = 0, 
                      delta_t = 0, h2o_fixed = "vapor pressure",
                      atmosphere = "tropical", clouds = "none", 
                      altitude_km = 70, looking = "down")
    
          # Read the MODTRAN results into R
          results = read_modtran(file_name)

          p = plot_modtran(file_name, descr = str_c(p_methane, " ppm methane"))
          print(p)
          # Create a data tibble with columns for the methane concentration
          # and I out, and append it to the end of the tibble methane_data
          df = tibble(methane = results$ch4, i_out = results$i_out)
          methane_data = bind_rows(methane_data, df)
        }
```

Now that we have completed all the model runs, we can analyze the data.
First, plot I~out~ for each concentration of methane:

```{r plot_methane_saturation, include=TRUE}
  plot_0 = ggplot(methane_data, aes(x = methane, y = i_out)) +
  geom_line(size = 1) + 
  geom_point(size = 2) +
  labs(x = "Methane concentration (ppm)", 
       y = expression(I[out]~(W/m^2)),
       title = expression(paste(I[out], " versus methane concentration")))

  print(plot_0)
```

This plot shows saturation in action: At small concentrations of methane,
a small amount of added methane produces a large decrease in I~out~, but
when methane concentrations are larger, even large additions of methane
have a much smaller effect on I~out~ than the small additions did when 
concentrations were low.

However, to examine saturation, it's often much more useful to plot
I~out~ against the logarithm of the concentration, as we show below
by plotting the concentrations on a logarithmic axis. 

A logarithmic axis represents the intervals for doublings as the same 
size on the axis, whether they represent a doubling from 1 to 2 ppm or 
from 1000 to 2000 ppm.

```{r plot_methane_saturation_log_scale, include=TRUE}
  plot_1 = ggplot(methane_data, aes(x = methane, y = i_out)) +
  geom_line(size = 1) + 
  geom_point(size = 2) +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50, 100, 200, 500, 1000)) +
  labs(x = "Methane concentration (ppm)", 
       y = expression(I[out]~(W/m^2)),
       title = expression(paste(I[out], 
                                " versus methane concentration (log scale)")))

  print(plot_1)
```

When we look at I~out~ versus the logarithm of the concentration, we see
that at small concentrations, successive doublings produce steeper and
steeper slopes in the change of I~out~, but as concentrations get large
the slopes change by less and less from one doubling to the next and 
I~out~ begins to approach a straight line instead of a curve.

In order to better estimate where I~out~ stops curving and becomes straight,
we plot the change in I~out~ from one
concentration to the next (remember that we're doubling the concentration
each time). 

Use the `lag` function to calculate how `i_out` changes from one row
to the next:

```{r calc_i_out_change, include = TRUE}
methane_data = methane_data %>% mutate(change = i_out - lag(i_out))
```


When I~out~ starts to follow a straight line when it's plotted
against the logarithm of the concentration:

```{r plot_methane_intensity_change, include=TRUE}
  plot_2 = ggplot(methane_data, aes(x = methane, y = change)) +
  geom_line(size = 1) + 
  geom_point(size = 2) +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50, 100, 200, 500, 1000)) +
  labs(x = "Methane concentration (ppm)", 
       y = expression(Delta * I[out]~(W/m^2)),
       title = expression(paste("Change in ", I[out], 
                                " between successive doublings")))

  print(plot_2)
```

At small concentrations, we see the change in I~out~ getting more and 
more negative, but at larger concentrations, as saturation kicks in, 
we start to see the change in I~out~ flattening out. 

We can identify band saturation with the first concentration at which the 
change in I~out~ is the same for that doubling and for the next.
This would be around 200 ppm.

You notice that the flattening isn't perfect or absolute: the changes
flatten out from 200--400 ppm, but then become even smaller at 800 ppm.
This is because many molecules, such as methane, have multiple absorption
bands, and the different bands saturate at different concentrations.

We generally say that absorption begins to saturate when the successive
changes in I~out~ flatten out for the first time.

**Answer:** 
The graph of the change in I~out~ for successive doublings of methane concentration
flattens out at around 200 ppm, meaning that we can estimate that this is 
roughly where band saturation occurs.
  
c) Would a doubling of methane have as great an impact on the heat balance as a 
   doubling of CO~2~?
   
**Answer:**
We already have a baseline run of MODTRAN from part (a), 
with 400 ppm CO~2~ and 1.7 ppm methane, so now we need to run MODTRAN
for 800 ppm CO~2~ and 1.7 ppm methane, and for 
400 ppm CO~2~ and 3.4 ppm methane.

```{r ex_4_1_c, include=TRUE, message=FALSE}
run_modtran(file.path(data_dir, "modtran_double_co2.txt"), co2 = 800, ch4 = 1.7, 
            atmosphere = 'tropical')
run_modtran(file.path(data_dir, "modtran_double_ch4.txt"), co2 = 400, ch4 = 2 * 1.7, 
            atmosphere = 'tropical')
double_co2 = read_modtran(file.path(data_dir, "modtran_double_co2.txt"))
double_methane = read_modtran(file.path(data_dir, "modtran_double_ch4.txt"))
```

Now compare I~out~ and calculate the change of I~out~ from the baseline
when we double CO~2~ and when we double methane:

```{r ex_4_1_c_calc_i_out, include=TRUE, message=FALSE}
i_2x_co2 = double_co2$i_out
i_2x_methane = double_methane$i_out

delta_co2_2x = i_baseline - i_2x_co2
delta_methane_2x = i_baseline - i_2x_methane
```

Doubling CO~2~ reduces I~out~ by 
`r format_md(delta_co2_2x, digits = 2)` W/m^2^.
Doubling methane reduces I~out~ by 
`r format_md(delta_methane_2x, digits = 2)` W/m^2^, so 
the effect of doubling CO~2~ on I~out~ is
`r format_md(delta_co2_2x / delta_methane_2x, digits = 2)`
times greater than doubling methane, 
and doubling CO~2~ will have a much greater effect on temperature.

d) What is the "equivalent CO~2~" of doubling atmospheric methane? That is to say,
   how many ppm of CO~2~ would lead to the same change in outgoing IR radiation
   energy flux as doubling methane? What is the ratio of ppm CO~2~ change to 
   ppm methane change?

**Answer:**
From part (c), we know that doubling methane reduces I~out~ by
`r format_md(delta_methane_2x, digits = 2)` W/m^2^.

Part (a) showed us that the effect on I~out~ of adding 10 ppm of methane
is `r format_md(ratio_10_ppm, digits = 1)` times as great as adding 
10 ppm of CO~2~, so we might guess that doubling methane (adding 1.7 ppm) 
would be equivalent to adding 
`r format_md(ratio_10_ppm * 1.7, digits = 1)` ppm of CO~2~.

Let's test this guess with MODTRAN:

```{r ex_4_1_d_guess, include=TRUE, message=FALSE}
delta_co2_guess = ratio_10_ppm * 1.7

run_modtran(file.path(data_dir, "modtran_guess.txt"), atmosphere = 'tropical',
            co2 = 400 + delta_co2_guess, ch4 = 1.7)
modtran_guess_co2 = read_modtran(file.path(data_dir, "modtran_guess.txt"))

i_guess = modtran_guess_co2$i_out

delta_i_guess = i_baseline - i_guess
```

So we see that our guess of increasing CO~2~ by 
`r format_md(delta_co2_guess, digits = 1)` ppm 
changed I~out~ by 
`r format_md(delta_i_guess, digits = 2)` W/m^2^, 
so this was not nearly enough of a change in CO~2~ 
to match a doubling of methane.

```{r ex_4_1_manual_equiv, include=TRUE, message=FALSE}
equiv_co2 = 85
```

Next, open up the web-based version of MODTRAN and manually adjust
the CO~2~ concentration until I~out~ matches I~out~ for doubled 
methane. If you do this, you will find that adding
`r format_md(equiv_co2, digits = 1)` ppm of CO~2~
matches doubling methane:

```{r ex_4_1_manual, include=TRUE, message=FALSE}
run_modtran(file.path(data_dir, "modtran_equiv_co2.txt"), 
            atmosphere = 'tropical', co2 = 400 + equiv_co2, ch4 = 1.7)

modtran_equiv_co2 = read_modtran(file.path(data_dir, "modtran_equiv_co2.txt"))

i_equiv = modtran_equiv_co2$i_out

delta_equiv = i_baseline - i_equiv
```

Adding `r format_md(equiv_co2, digits = 1)` ppm CO~2~ 
changes I~out~ by
`r format_md(delta_equiv, digits = 2)` W/m^2^, which
matches what we measured for doubling methane.

Thus, doubling methane is equivalent to increasing CO~2~ by
`r equiv_co2` ppm.

## Exercise 4.2: CO~2~ (Graduate students only)

a) Is the direct effect of increasing CO~2~ on the energy output at the top of
   the atmosphere larger in high latitudes or in the tropics?

    For each atmosphere, first record $I_{\text{out}}$ with CO~2~ at 400 ppm
    and then record the change when you increase CO~2~ to 800 ppm.


**Answer:**
First, run MODTRAN for the different configurations:

```{r ex_4_2_a, include=TRUE, message=FALSE}
# We have already calculated these numbers for tropical atmosphere.
# Now let's do the same for midlatitude summer and subarctic summer:
run_modtran(file.path(data_dir, "midlat_summer_baseline.txt"), co2 = 400,
            atmosphere = 'midlatitude summer') 
run_modtran(file.path(data_dir, "midlat_summer_2x_co2.txt"), co2 = 800,
            atmosphere = 'midlatitude summer') 
run_modtran(file.path(data_dir, "subarctic_summer_baseline.txt"), co2 = 400,
            atmosphere = 'subarctic summer') 
run_modtran(file.path(data_dir, "subarctic_summer_2x_co2.txt"), co2 = 800,
            atmosphere = 'subarctic summer')

midlat_baseline = read_modtran(file.path(data_dir, 
                                         "midlat_summer_baseline.txt"))
midlat_double = read_modtran(file.path(data_dir, 
                                       "midlat_summer_2x_co2.txt"))

i_midlat_baseline = midlat_baseline$i_out
i_midlat_double = midlat_double$i_out

delta_midlat =  i_midlat_baseline - i_midlat_double

subarctic_baseline = read_modtran(file.path(data_dir, 
                                            "subarctic_summer_baseline.txt"))
subarctic_double = read_modtran(file.path(data_dir, 
                                          "subarctic_summer_2x_co2.txt"))

i_subarctic_baseline = subarctic_baseline$i_out
i_subarctic_double = subarctic_double$i_out

delta_subarctic = i_subarctic_baseline - i_subarctic_double
```

Here are the results:

* Tropical: Change in I~out~ from doubling CO~2~ is 
  `r format_md(delta_co2_2x, digits = 2)` W/m^2^.

* Midlatitude: Change in I~out~ from doubling CO~2~ is 
  `r format_md(delta_midlat, digits = 2)` W/m^2^.

* Subarctic: Change in I~out~ from doubling CO~2~ is 
  `r format_md(delta_subarctic, digits = 2)` W/m^2^.

So the effect of doubling CO~2~ is strongest in the tropics and gets weaker the
farther toward the poles you go. 

**Note:** This is what happens without feedbacks. 
If we include the important feedbacks in the climate system,
the effect of doubling CO~2~ is much greater near the poles
than in the tropics.

b) Set pCO~2~ to an absurdly high value of 10,000 ppm. You will see a spike
   in the CO~2~ absorption band. What temperature is this light coming from? 
   Where in the atmosphere do you think this comes from?

    Now turn on clouds and run the model again. Explain what you see.
    Why are night-time temperatures warmer when there are clouds?

**Answer:** As we add CO~2~, the skin height rises. 

When the emission at a
certain wavelength flattens out, that corresponds to the skin height for
that wavelength reaching the tropopause, so as the skin height rises
the temperature at the skin height remains constant because it's in the 
region of the lower stratosphere where the environmental lapse rate is zero.

Eventually, the skin height rises above that part of the stratosphere into the
region where the temperature starts rising with increasing altitude (i.e.,
where the environmental lapse rate is negative). 

For wavelengths where the 
skin height is in this part of the stratosphere, there will be a spike of 
increased longwave emissions.

**Details:** 

Students don't need to do this, but this makes a good illustration of exactly
where in the atmosphere, the spike comes from:

```{r ex_4_2_spectra, include=TRUE, message=FALSE}
pco2 = 1.0E+4 # 10,000 ppm

for (alt in c(10, 20, 30, 40, 50, 60, 70)) {
  fname = str_c(file.path(data_dir, "10k_co2_alt_"), alt, '.txt')
  run_modtran(fname, co2 = pco2, altitude = alt)
  p = plot_modtran(fname)
  print(p)
}
```

Now let's look at the temperature profile:

```{r ex_4_2_profile, include=TRUE, message=FALSE, warning = FALSE}
profile = read_modtran_profile(file.path(data_dir, "10k_co2_alt_70.txt"))

ggplot(profile, aes(x = T, y = Z)) +
  geom_path(size = 1) +
  scale_y_continuous(breaks = seq(0,70,10), limits = c(0,70)) +
  labs(x = "Temperature (Kelvin)", y = "Altitude (km)")
```


Now let's run the model with the altitude set to 70 km and stratus clouds
turned on:

```{r ex_4_2_b_clouds, include=TRUE, message=FALSE}
pco2 = 1.0E+4 # 10,000 ppm

run_modtran(file.path(data_dir, "modtran_10k_co2_no_clouds.txt"), co2 = pco2,
            clouds = 'none')

run_modtran(file.path(data_dir, "modtran_10k_co2_clouds.txt"), co2 = pco2,
            clouds = 'altostratus')

no_clouds = read_modtran(file.path(data_dir, "modtran_10k_co2_no_clouds.txt"))
clouds = read_modtran(file.path(data_dir, "modtran_10k_co2_clouds.txt"))

i_no_clouds = no_clouds$i_out
i_clouds = clouds$i_out

p_no_clouds = plot_modtran(file.path(data_dir, "modtran_10k_co2_no_clouds.txt"))
print(p_no_clouds)

p_clouds = plot_modtran(file.path(data_dir, "modtran_10k_co2_clouds.txt"))
print(p_clouds)
```

Without clouds, the outgoing heat from longwave radiation is 
I~out~ = `r format_md(i_no_clouds, digits = 2)` and with clouds, it's 
I~out~ = `r format_md(i_clouds, digits = 2)`.

In the plots of the spectra, notice how the emissions in the infrared window
from 800--1200 cm^-1^ decrease when middle-level clouds (altostratus) are
added.

## Exercise 4.3: Water vapor

Our theory of climate presumes that an increase in the temperature at ground
level will lead to an increase in the outgoing IR energy flux at the top of the
atmosphere.


a) How much extra outgoing IR would you get by raising the temperature of the 
   ground by 5&deg;C? What effect does the ground temperature have on the 
   shape of the outgoing IR spectrum and why?

**Answer:**
Run MODTRAN with `delta_t` set to 5 Kelvin:

```{r ex_4_3_a, include=TRUE, message=FALSE}
run_modtran(file.path(data_dir, "modtran_plus_5k.txt"), delta_t = 5, h2o_fixed = 'vapor pressure')

modtran_5k = read_modtran(file.path(data_dir, "modtran_plus_5k.txt"))

i_5k = modtran_5k$i_out

delta_5k =  i_5k - i_baseline
```

Raising the ground temperature by 5K raises I~out~ by 
`r format_md(delta_5k, digits = 2)` W/m^2^.

Next, plot the baseline spectrum and the spectrum for the
warmer surface so we can compare them:

```{r plot_delta_5k, include=TRUE, message=FALSE}
p_baseline = plot_modtran(file.path(data_dir, "modtran_baseline.txt"), 
                          descr = "Baseline")
print(p_baseline)

p_5k = plot_modtran(file.path(data_dir, "modtran_plus_5k.txt"), 
                    descr = "Temperature increased by 5K")
print(p_5k)
```

The whole spectrum becomes brighter (warmer), but you can see a 
greater increase in the wavelengths corresponding to the
infrared window (800--1200 cm^-1^).

You can also see changes in the main CO~2~ emissions peak around 650 cm^-1^,
but not as much as in the window region.

b) More water can evaporate into warm air than into cool air. Change the
   model settings to hold the water vapor at constant relative humidity 
   rather than constant vapor pressure (the default), calculate the change
   in outgoing IR energy flux for a 5&deg;C temperature increase.
   Is it higher or lower? Does water vapor make the Earth more sensitive to
   CO~2~ increases or less sensitive?

**Answer:**
Run MODTRAN with relative humidity fixed and compare I~out~ to
what we saw in part (a) when we held vapor pressure fixed.

```{r ex_4_3_b, include=TRUE, message=FALSE}
run_modtran(file.path(data_dir, "modtran_plus_5k_humidity.txt"), 
            delta_t = 5, h2o_fixed = 'relative humidity')

modtran_5k_humidity = 
  read_modtran(file.path(data_dir, "modtran_plus_5k_humidity.txt"))

i_5k_humidity = modtran_5k_humidity$i_out
delta_5k_humidity = i_5k_humidity - i_baseline
```

When we raised the surface temperature by 5K with 
vapor pressure constant, I~out~ changed by 
`r format_md(delta_5k, digits=2)`.
When we raise the surface temperature by 5K with 
relative humidity constant, I~out changes by
`r format_md(delta_5k_humidity, digits=2)`, which is
about 
`r format_md(100 * delta_5k_humidity / delta_5k, digits = 0)` 
percent of what it was with the water vapor pressure held constant.

Constant relative humidity reduces the change in outgoing longwave radiation
for the same in temperature, so the temperature would have to rise higher to
compensate for the decrease in I~out~ when we increase CO~2~. 

A given change in temperature produces a smaller change in I~out~ when 
relative humidity is constant (i.e., when the water vapor feedback is 
active) than when vapor pressure is held constant (i.e., when the
water vapor feedback is disabled).

When a forcing is applied (e.g., increasing greenhouse gas concentrations),
it changes I~out~, and the surface temperature must change enough to 
bring I~out~ back to its original value to balance the heat flow.

This means
that water vapor makes the earth _more_ sensitive to changes in CO~2~.

Now, let's plot the spectrum for a 5K increase in surface temperature
with constant relative humidity.

```{r humidity_plot, include=TRUE, message=FALSE}
p_humidity = plot_modtran(file.path(data_dir, "modtran_plus_5k_humidity.txt"),
                          descr = "Temp. increased by 5K (const. rel. hum.)")
print(p_humidity)
```

c) Now see this effect in another way. 

    * Starting from the default base case, record the total outgoing 
      IR flux. 

    * Now double pCO2. The temperature in the model stays the
      same (that's how the model is written), but the outgoing IR flux
      goes down.

    * Using constant water vapor pressure, adjust the temperature offset
      until you get the original IR flux back again. Record the change in
      temperature
    
    * Now repeat the exercise, but holding the relative humidity fixed
      instead of the water vapor pressure.
    
    * The ratio of the warming when you hold relative humidity fixed
      to the warming when you hold water vapor pressure fixed is the 
      feedback factor for water vapor. What is it?

**Answer:** 
Under baseline conditions (400 ppm CO~2~), 
I~out~ = `r format_md(i_baseline, digits = 2)` W/m^2^.

```{r ex_4_3_c_interactive, include=TRUE, message=FALSE}
#
# Figure out the right delta_t to use by interactively playing with the
# web-based interface to MODTRAN. Then insert the values here.
#
delta_t_vapor_pressure = 0.76
delta_t_humidity = 1.21
```

After playing with the interactive web-based MODTRAN, we find that
after doubling CO~2~ with constant water vapor pressure,
a warming of
`r delta_t_vapor_pressure` Kelvin restores I~out~ to
its original value for 400 ppm.

After doubling CO~2~ with constant relative humidity, it takes a warming of
`r delta_t_humidity` Kelvin to restore I~out~ to its original value.

The ratio of warming for constant relative humidity versus constant water vapor
pressure is 
`r format_md(delta_t_humidity / delta_t_vapor_pressure, digits = 1)`,
so the feedback factor for water vapor in the tropics is 
_f_ = `r format_md(delta_t_humidity / delta_t_vapor_pressure, digits = 1)`.

Below, we show that these changes in temperature restore the original
(baseline) I~out~ with CO~2~ at 800 ppm:

```{r ex_4_3_c, include=TRUE, message=FALSE}
#
# Figure out the right delta_t to use by interactively playing with the
# web-based interface to MODTRAN. Then insert the values here.
#
run_modtran(file.path(data_dir, "warming_vapor_pressure.txt"), 
            co2 = 800, delta_t = delta_t_vapor_pressure, h2o_fixed = 'vapor pressure')

run_modtran(file.path(data_dir, "warming_humidity.txt"), 
            co2 = 800, delta_t = delta_t_humidity, h2o_fixed = 'relative humidity')

warming_vapor_pressure = read_modtran(file.path(data_dir, "warming_vapor_pressure.txt"))
warming_humidity = read_modtran(file.path(data_dir, "warming_humidity.txt"))

i_vapor_pressure = warming_vapor_pressure$i_out
i_humnidity = warming_humidity$i_out
```

Under default conditions, 
I~out~ = `r format_md(i_baseline, digits = 2)` W/m^2^.

With doubled CO~2~ and constant water vapor pressure,
raising the surface temperature by 
`r delta_t_vapor_pressure` Kelvin restores I~out~ to
`r format_md(i_vapor_pressure, digits = 2)` W/m^2^.

With doubled CO~2~ and constant relative humidity, 
raiding the surface temperature by 
`r delta_t_humidity` Kelvin restores I~out~ to
`r format_md(i_humnidity, digits = 2)` W/m^2^.

# Chapter 5 Exercise

## Exercise 5.2: Skin Height

a) Run the MODTRAN model in using the "Tropical" atmosphere, without clouds, and with
   present-day pCO~2~ (400 ppm). Use the ground temperature reported by the model to calculate
   $\varepsilon \sigma T_{\text{ground}}^4$, the heat flux emitted by the ground.
   Assume $\varepsilon = 1$, and I have already provided the value of the 
   Stefan-Boltzmann constant $\sigma$, as the R variable `sigma_sb`,
   which equals `r format_md(sigma_sb, digits = 3, format = "scientific")`.
   (I defined it in the script "utils.R", which I loaded in the "setup" chunk
   in the RMarkdown document).

    Next, look at the outgoing heat flux at the top of the atmosphere (70 km) 
    reported by the MODTRAN model. Is it greater or less than the heat flux
    that you calculated was emitted by the ground?

**Answer:** 
Use the Stefan-Boltzmann law to calculate I~up,ground~:

```{r ex_5_2_a, include=TRUE, message=FALSE}
T_ground = baseline$t_ground

i_up_ground = sigma_sb * T_ground^4
i_up_skin = baseline$i_out
```

The ground temperature is `r format_md(T_ground, digits = 1)` Kelvin,
so the Stefan-Boltzmann equation tells us that the ground emits
I~up,ground~ = `r format_md(i_up_ground, digits = 2)` W/m^2^ of longwave
radiation.

MODTRAN calculates that at the top of the atmosphere, there is
I~out~ = `r format_md(i_up_skin, digits = 2)` W/m^2^ of longwave radiation
going out to space, which is considerably less than what we calculated was
emitted by the ground.

b) Use the outgoing heat flux at the top of the atmosphere to calcuate the
   skin temperature (use the equation 
   $I_{\text{out}} = \varepsilon \sigma T_{\text{skin}}^4)$).
   What is the skin temperature, and how does it compare to the ground 
   temperature and the temperature at the tropopause, as reported by the 
   MODTRAN model?
   
    Assuming an environmental lapse rate of 6K/km, and using the 
    skin temperature that you calculated above, and the ground temperature
    from the model, what altitude would you expect the skin height to be?

**Answer:**
The Stefan-Boltzmann law tells us that
$$I = \varepsilon\sigma T^4,$$
so we can do a little algebra to figure out that
$$T = \sqrt[4]{\frac{I}{\varepsilon\sigma}}.$$
We can use this equation to calcualte the effective
skin temperature from I~out~ at the top of the atmosphere.

```{r ex_5_2_b_skin_temp, include=TRUE, message=FALSE}
T_skin = (i_up_skin / sigma_sb)^0.25 # from Stefan-Boltzmann law
```

T~skin~ = `r format_md(T_skin, digits = 2)` K.

Next, use the lapse rate and the difference between the
skin temperature and the ground temperature in order
to calculate the skin height:
$$h_{\text{skin}} = 
\frac{T_{\text{ground}} - T_{\text{skin}}}{\text{environmental lapse}}$$

```{r ex_5_2_b, include=TRUE, message=FALSE}
env_lapse = 6 # Kelvin per kilometer

h_skin = (T_ground - T_skin) / env_lapse

T_tropopause = baseline$t_tropo
h_tropopause = baseline$h_tropo
```

We find that h~skin~ = `r format_md(h_skin, digits = 2)` km.

According to MODTRAN, 
the ground temperature is `r format_md(T_ground, digits = 3)` K,
the temperature at the tropopause is 
`r format_md(T_tropopause, digits = 3)` K, 
and the height of the 
tropopause is 
`r format_md(h_tropopause, digits = 2)` km, 
so our estimate suggests that the skin height in the tropics
is well below the tropopause, and the skin temperature is considerably
warmer than the tropopause, but considerably colder than the ground
temperature.

c) Double the CO~2~ concentration and run MODTRAN again. Do not adjust the
   ground temperature. Repeat the calculations from (b) of the skin
   temperature and the estimated skin height.

    What is the new skin temperature? What is the new skin height?

**Answer:**
In part exercise 4.1 (c) we measured I~out~ for doubled CO~2~ 
with no surface temperature change, so we can use this for 
I~out~ from the skin-height with doubled CO~2~:

```{r ex_5_2_c, include=TRUE}
i_skin_2 = i_2x_co2
```

I~skin~ = `r format_md(i_skin_2, digits = 2)` W/^2^

Next, we use the Stefan-Boltzmann law to calculate the skin temperature, 
just as we did in part (b):

```{r ex_5_2_c_t_skin, include=TRUE}
T_skin_2 = (i_skin_2 / sigma_sb)^0.25
```

T~skin~ = `r format_md(T_skin_2, digits = 2)` K.
Now we can use the lapse rate to calculate the skin height:

```{r ex_5_2_c_h_skin, include=TRUE}
h_skin_2 = (T_ground - T_skin_2) / env_lapse
```

The skin height is 
`r format_md(h_skin_2, digits = 1)` km,
which is
`r format_md(h_skin_2 - h_skin, digits = 1)` km 
higher with doubled CO~2~ than for the baseline (current conditions).

d) Put the CO~2~ back to today's value, but add cirrus clouds, using the
   "standard cirrus" value for the clouds. Repeat the calculations from (b) of
   the skin temperature and the skin height.
   
    What is the new skin temperature? What is the new skin height?
    Did the clouds or the doubled CO~2~ have a greater effect on the
    skin height?

**Answer:**
Run MODTRAN with cirrus clouds and compare to the baseline conditions:

```{r ex_5_2_d, include=TRUE, message=FALSE}
run_modtran(file.path(data_dir, "modtran_cirrus.txt"), clouds = 'standard cirrus')

cirrus = read_modtran(file.path(data_dir, "modtran_cirrus.txt"))

i_cirrus = cirrus$i_out

T_skin_cirrus = (i_cirrus / sigma_sb)^0.25

h_skin_cirrus = (T_ground - T_skin_cirrus) / env_lapse
```

With cirrus clouds (using the standard cirrus model),
the skin temperature is
`r format_md(T_skin_cirrus, digits = 2)` Kelvin 
and the skin height is
`r format_md(h_skin_cirrus, digits = 2)` kilometers.
which is
`r format_md(h_skin_cirrus - h_skin, digits = 1)` km higher than for 
the baseline (current conditions) and
`r format_md(h_skin_cirrus - h_skin_2, digits = 1)` km higher than for 
doubled-CO~2~ conditions.

Cirrus clouds had much bigger effect than doubling CO~2~.
